The Community Land Model, version 5.1 \\ One-at-a-time Parameter Perturbation Ensemble

### 2024-05-22 ms/ms.tex

\section{Introduction}
Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within the terrestrial components of Earth System Models. Understanding how these processes respond to and influence CO$_2$ concentrations is a critical facet of climate change research. Our uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}. Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}. Our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns. However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development.

Model inter-comparison projects (MIPs) are a major component of the model development cycle, and have been the primary means by which model projection uncertainty is assessed \cite{henderson-sellers1995,pitman1999,wood1998,schlosser2000,eyring2016,friedlingstein2022}. While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}. In most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the high cost of each simulation, e.g. the TRENDY land model intercomparison or the Coupled Model Intercomparison Project \cite{friedlingstein2022,eyring2016}. Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.

@@

The process of parameter estimation for land models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.). The number of model parameters required to represent all the relevant processes is large, and model speeds are relatively slow, presenting a major barrier to robust objective calibration, model interpretation and development \cite{fisher2020,dagon2020}.
Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}. It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge.
Our work here tries to alleviate each of these challenges in turn, at least within the context of the CLM-BGC model.
While we are not yet at the point of automated calibration, we present herein several innovations that will serve to make an automated calibration system more technically feasible.

In this project we create a large PPE with the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC). We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters. Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
Our goal in this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes. 
In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, describe how it was produced, and survey some potential applications.
The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.

@@

To manage computational cost we leveraged the Semi-Analytic spin-up mode (i.e., SASU) recently implemented within CLM5 \cite{lu2020} (Liao et al. 2023). This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude (Liao et al. 2023, Luo et al. 2022). Our spin-up protocol featured 20 years in Accelerated Decomposition (AD) spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e. ND), cycling over a ten-year forcing dataset (described below, Liao et al. 2023). This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time. This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth) in very few points at high latitude. Certain inferences about deep soil carbon from this ensemble are therefore subject to uncertainty due to spin-up concerns.

@@

Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis. We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}). The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}. We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings. Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}; AF2095 and AF1855). We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables. Our future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in order to allow comparison with the existing CESM2 large ensemble. Our software infrastructure has greatly reduced the effort and computational cost associated with ensemble generation, such that it would be easy to add additional ensembles, to explore other emission scenarios or climate regimes (e.g. Last Glacial Maximum).

@@

Given the challenging size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the need to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
Combining the SASU spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for our experiment within 140 years.

@@

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Given more clusters, the sparsegrid will generally provide a better approximation of output from the full grid. As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz). We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
\begin{figure}[h]
\centering
\includegraphics[width=25pc]{../figs/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

Overall, we found substantial impacts of parametric uncertainty on model behaviour, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}). One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death. A perturbation of $\pm$20\% likely exceeds the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils. Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, including unforeseen tipping points, brittle parameterizations,  and/or bugs.

A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}) for any given variable or metric. For example, the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig \ref{supp:fcev}). In general,most of the parameter perturbations had a relatively small impact on any given output variable. As such, the effective parameter space dimensionality for any given CLM process is likely much smaller than would be expected from having 211 independent parameters, reflecting the fact that land models like CLM comprise representations of numerous interconnected domains. While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.

@@

We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}. This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control responses to individual forcing variables (CO2, nitrogen deposition) and climate.  In the case of evapotranspiration (ET), for example, the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}). 
ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
Besides \textit{kmax}, all of the parameters that appear in the top eight parameters for ET response to future climate govern plant acclimation to temperature. 
These parameters, while not especially influential on present-day ET, would be among the most influential in governing the trend in ET going forward, illustrating the potential limitations of calibration to ambient conditions.

The PPE is likewise useful for determining which parameters influence which model processes and where. For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}). Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome. The most influential parameters for leaf area index globally include parameters that were important in each of these three biomes.

@@

Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data via approximately 2000 plots (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).

The diagnostics website includes ranking plots (as in Figure \ref{fig:lai}) and maps of parameter effects (as in Figure \ref{fig:panel}). But these plots are repeated across a combination of model output variables and model parameters. Likewise figures are repeated for each of the various forcing scenarios.

>>>>

\section{Introduction}
Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within the terrestrial components of Earth System Models. Understanding how these processes respond to and influence CO$_2$ concentrations is a critical facet of climate change research. Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}. Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}. Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns. However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development.

Model inter-comparison projects (MIPs) are a major component of the model development cycle, and have been the primary means by which model projection uncertainty is assessed \cite{henderson-sellers1995,pitman1999,wood1998,schlosser2000,eyring2016,friedlingstein2022}. While MIPs have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}. In most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the high cost of each simulation, e.g. the TRENDY land model intercomparison or the Coupled Model Intercomparison Project \cite{friedlingstein2022,eyring2016}. Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.

@@

The process of parameter estimation for the complex land models typically embedded within Earth System Model is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.). The number of model parameters required to represent all the relevant processes is large, and model simulations are relatively expensive, presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}. It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge.
Our work here tries to alleviate each of these challenges in turn, at least within the context of one land model.
While we are not yet at the point of automated calibration, we present herein several innovations that will serve to make an automated calibration system more technically feasible.

In this project we create a large PPE with the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC). We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters. Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
The goal of this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes. 
In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, describe how it was produced, and survey some potential applications.
The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.

@@

To manage computational cost we leveraged the Semi-Analytic spin-up mode (i.e., SASU) recently implemented within CLM5 \cite{lu2020} (Liao et al. 2023). This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude (Liao et al. 2023, Luo et al. 2022). The spin-up protocol featured 20 years in Accelerated Decomposition (AD) spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e. ND), cycling over a ten-year forcing dataset (described below, Liao et al. 2023). This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time. This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth) in very few points at high latitude. Certain inferences about deep soil carbon from this ensemble are therefore subject to uncertainty due to spin-up concerns.

@@

Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis. We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}). The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}. We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings. Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}; AF2095 and AF1855). We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables. The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in order to allow comparison with the existing CESM2 large ensemble. Our software infrastructure has greatly reduced the effort and computational cost associated with ensemble generation, such that it would be easy to add additional ensembles, to explore other emission scenarios or climate regimes (e.g. Last Glacial Maximum).

@@

Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the desire to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
Combining the SASU spinup approach with the sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5 simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for this experiment within 140 years.

@@

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Given more clusters, the sparsegrid will generally provide a better approximation of output from the full grid. As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters (Figure \ref{fig:sg}).
Performance continues to improve with added clusters, but with marginal returns.
We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters. Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations (Supp Figure S2). We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording us sufficient simulations to perform the full experiment within our computational budget.
\begin{figure}[h]
\centering
\includegraphics[width=25pc]{../figs/main/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

This model configuration allowed for over two thousand simulations across six forcing scenarios.
Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}). In the control scenario, gross primary productivity (GPP) ranged from 103 to 158 PgC, and net ecosystem productivity (NEP) ranged from 1.7 to 3.8 PgC. NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC. One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death. A perturbation of $\pm$20\% may exceed the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils. Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, including unforeseen tipping points, brittle parameterizations,  and/or bugs.

A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}) for any given variable or metric. For example, just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3). Conversely, GPP, which is a significantly more complicated process, requires 20 parameters to explain 90\% of the control ensemble variance. In general, most of the parameter perturbations had a relatively small impact on any given output variable. Though this experiment encompasses 211 parameters, no single variable responds to all 211 parameters. In fact, the effective parameter space dimensionality for any given output variable tends to be much smaller than 211. While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.

@@

This PPE is especially useful for determining which parameters influence which model processes and where. For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}). Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome. The most influential parameters for leaf area index globally included parameters that were important in each of these three biomes.

We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}. This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control the response to forcing (CO2, nitrogen deposition) and climate.
For example, the parameters that control present-day NEP differ  from the parameters that control NEP in the high CO$_2$ scenario (Figure \ref{fig:nep}a,b). NEP increases by approximately 50\% at 867 ppm with the default parameters, but can actually decrease with certain parameter settings (Figure \ref{fig:nep}c). One key parameter is \textit{tpuse\_sf}, which is a scalar perturbation factor influencing the triose phosphate limitations on photosynthesis. Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations (cite Kumarathunge et al). In CLM5.1, this parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential with high CO$_2$.

The future climate experiment is 3K(zqz) warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S4). This led to an increase in evapotranspiration of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f). While the parameters that most strongly affect global average ET in both the control and future climates are generally the same (Figure \ref{fig:nep}d,e), a distinct set of parameters govern the change in ET due to warming (Figure \ref{fig:nep}f). While the control simulation is primarily influenced by hydrology and stomatal conductance parameters, the response of ET to warming is largely controlled by photosynthesis acclimation parameters (i.e. \textit{tpuse\_sf}, \textit{kcha}, \textit{vcmaxha}, \textit{vcmaxhd}, \textit{lmrhd}, and \textit{cpha}). The scale of ET response may be inflated, given that these are land-only experiments, and an interactive atmosphere could modulate the ET response.

@@

Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data via approximately 2000 plots (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}). The diagnostics website includes ranking plots (as in Figure \ref{fig:lai}) and maps of parameter effects (as in Figure \ref{fig:panel}). But these plots are repeated across a combination of model output variables and model parameters. Likewise figures are repeated for each of the various forcing scenarios.
